---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- #' # -->
<!-- #' TODO: -->
<!-- #' sk_plot_semi needs an explanation for getting vg -->
<!-- #' sk.plot refers to graphics::... when it should be grDevices -->

# snapKrig

An R package for fast spatial analysis and kriging on grids

<!-- badges: start -->

[![CRAN status](https://www.r-pkg.org/badges/version/snapKrig)](https://CRAN.R-project.org/package=snapKrig)
[![Lifecycle: stable](https://img.shields.io/badge/lifecycle-stable-brightgreen.svg)](https://lifecycle.r-lib.org/articles/stages.html#stable)
[![](https://cranlogs.r-pkg.org/badges/snapKrig)](https://cran.r-project.org/package=snapKrig)

<!-- badges: end -->

snapKrig uses a computationally lean implementation of a 2-dimensional spatial correlation model for gridded data. By restricting to models with (separable) Kronecker covariance, the package can speed computations on certain likelihood and kriging problems by orders of magnitude compared to alternatives like [gstat](https://cran.r-project.org/package=gstat), [fields](https://cran.r-project.org/package=fields), [geoR](https://cran.r-project.org/package=geoR), [spatial](https://cran.r-project.org/package=spatial), and [LatticeKrig](https://cran.r-project.org/package=LatticeKrig).

Here are some benchmarking results for computation time to downscale a 32 x 43 raster onto grids of increasing size by ordinary kriging.

```{r preview, echo = FALSE, message = FALSE, warning = FALSE, fig.width=6, fig.height=2, dpi=300}
library(here)
library(ggplot2)

path_bench_result = here('rjarticle/data/bench_results.csv')

results_df = read.csv(path_bench_result, row.names=NULL)
results_df = results_df[results_df[['name']] == 'treed_1376',]
results_df[['is_complete']] = c('no', 'yes')[1 + as.integer(results_df[['complete']])]

# make a plotting data frame with single column for both times
n_case = nrow(results_df)
results_plot = results_df[rep(seq(n_case), 2),]
results_plot[['with_var']] = rep(c('yes', 'no'), each=n_case)
results_plot[['teval']] = results_plot[['teval_pred']]
results_plot[['teval']][seq(n_case)] = results_plot[['teval_both']][seq(n_case)]

# make the plot
ggplot(results_plot) +
  aes(x=n_out, y=teval, color=pkg, lty=with_var) +
  geom_point(size=1, pch=1) +
  geom_line(linewidth=0.5) +
  scale_linetype_manual(values = c(no='11', yes='solid')) +
  xlab('number of points predicted') + 
  ylab('prediction time (seconds)') +
  labs(color='R package',
       lty='with variance') +
  scale_x_log10(
   breaks = scales::trans_breaks('log10', function(x) 10^x),
   labels = scales::trans_format('log10', scales::math_format(10^.x))
  ) +
  scale_y_log10(
   breaks = scales::trans_breaks('log10', function(x) 10^x),
   labels = scales::trans_format('log10', scales::math_format(10^.x))
  ) +
  theme_bw() +
  theme(text=element_text(size=8),
        strip.text.x=element_text(face='bold'),
        strip.text.y=element_text(face='bold'))

```


code and instructions to reproduce these results can be found [here](https://github.com/deankoch/snapKrig/tree/master/rjarticle/data)

## Installation

snapKrig is on [CRAN](https://cran.r-project.org/package=snapKrig). Install it with

``` r
install.packages('snapKrig')
```

Check out the [introduction vignette](https://CRAN.R-project.org/package=snapKrig/vignettes/snapKrig_introduction.html) for a worked example with the Meuse soils data, or try the code below to get started right away. Some other code examples be found [here](https://github.com/deankoch/snapKrig/tree/master/examples).  We plan to publish a more detailed tutorial and benchmarking study in an [upcoming paper](https://github.com/deankoch/snapKrig/tree/master/rjarticle). 

## Example

To get started define an empty grid

```{r intro_empty}
library(snapKrig)

# simulate data on a rectangular grid
g_empty = sk(c(100, 200))
g_empty
```
Generate some random auto-correlated data for this grid

```{r intro_sim, out.width='75%', fig.dim=c(9, 5), fig.align='center'}
# set a random seed
set.seed(1234567)

# simulate data on a square grid
pars = sk_pars(g_empty)
g_sim = sk_sim(g_empty, pars)

# plot
plot(g_sim, main='snapKrig simulation')
g_sim
```

Downscale and filter noise by simple kriging

```{r intro_down}
# downscale and predict, timing computations
t_start = Sys.time()
g_down = sk_rescale(g_sim, down=10)
g_pred = sk_cmean(g_down, pars, X=0)

# print time elapsed in computation
t_end = Sys.time()
t_end - t_start
```


```{r intro_down_plot, out.width='75%', fig.dim=c(9, 5), fig.align='center'}
# plot grid
plot(g_pred, main='snapKrig prediction at 10X resolution')

# print summary
summary(g_pred)
```

# Technical Features

snapKrig's computational efficiency makes it useful in situations where interpolation or
down-scaling would pose a problem due to long computation times and/or high memory demands.
Such problems are common when working with geo-referenced data in earth sciences. snapKrig's features
include:

* anisotropic Gaussian processes on 2-dimensional regular grids for a choice of covariance kernels
* optimized computation of the likelihood function, generalized least squares, and kriging predictor/variance
* fast computations with missing data problems, and even faster in the complete data case 
* automated maximum likelihood model fitting and support for sample semi-variograms
* user friendly helper functions for raster down-scaling and point interpolation

`snapKrig` depends only on packages included by default in R (like `graphics` and `stats`), but supports raster and geometry classes from `sf` and `terra`. 


# History

An earlier implementation of snapKrig was called [pkern](https://github.com/deankoch/pkern). snapKrig
is a redesigned version that uses a more user-friendly S3 grid object class. 

pkern was an R implementation of some methods I developed in [my thesis](https://doi.org/10.7939/r3-91zn-v276)
for speeding up geostatistical computations involving large covariance matrices. The central idea is to model
spatial dependence using a separable 2-dimensional covariance kernel, defined as the product of two (1-dimensional)
univariate covariance kernels. This introduces special symmetries and structure in the covariance matrix, which are
exploited in this package for fast and memory-efficient computations.

I developed snapKrig to support a project to interpolate weather data, but the methods underlying snapKrig
are applicable more generally. See also [[1](https://doi.org/10.7939/r3-g6qb-bq70)],
where I use product kernels to study directions of anisotropy in a non-stationary random fields, and
[[2](https://doi.org/10.1007/s11538-021-00899-z), [3](https://doi.org/10.1098/rsif.2020.0434)], where I apply it to
fit a covariance structure, and to speed up calculations of dispersal kernel convolutions.
