---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# snapKrig

<!-- badges: start -->
<!-- badges: end -->

snapKrig is a computationally lean implementation of a 2-dimensional spatial correlation model for
gridded data. By restricting to models with (separable) Kronecker covariance, the package speeds
computations on certain likelihood and kriging problems (orders of magnitude difference) compared
to alternatives like `gstat`, `fields`, `geoR`, `spatial`, and `LatticeKrig`.

## Installation

You can install the development version of snapKrig from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("deankoch/snapKrig")
```

A CRAN release is expected soon.

## Example

Define an empty grid to get started

```{r intro_empty}
library(snapKrig)

# simulate data on a rectangular grid
g_empty = sk(c(100, 200))
g_empty
```
Generate some random auto-correlated data for this grid

```{r intro_sim, out.width='75%', fig.dim=c(9, 5), fig.align='center'}
# simulate data on a square grid and plot
pars = sk_pars(g_empty)
g_sim = sk_sim(g_empty, pars)
plot(g_sim, main='snapKrig simulation')
g_sim
```

Downscale and filter noise by simple kriging

```{r intro_down, out.width='75%', fig.dim=c(9, 5), fig.align='center'}
g_down = sk_rescale(g_sim, down=10)
g_pred = sk_cmean(g_down, pars, X=0)
plot(g_pred, main='snapKrig prediction at 10X resolution')
summary(g_pred)
```


# Technical Features

snapKrig's computational efficiency makes it useful in situations where interpolation or
down-scaling would pose a problem due to long computation times and/or high memory demands.
Such problems are common when working with geo-referenced data in earth sciences. Features include:

* anisotropic Gaussian processes on 2-dimensional regular grids for a choice of covariance kernels
* optimized computation of the likelihood function, generalized least squares, and kriging predictor/variance
* fast computations with missing data problems, and even faster in the complete data case 
* automated maximum likelihood model fitting and support for sample semi-variograms
* user friendly helper functions for raster down-scaling and point interpolation

`snapKrig` depends only on packages included by default in R (like `graphics` and `stats`), but supports 
raster and geometry classes from `sf` and `terra`. 


# History

An earlier incarnation of snapKrig was called [pkern](https://github.com/deankoch/pkern). snapKrig
is a redesigned version that uses a more user-friendly S3 grid object class. 

pkern was an R implementation of some methods I developed in [my thesis](https://doi.org/10.7939/r3-91zn-v276)
for speeding up geostatistical computations involving large covariance matrices. The central idea is to model
spatial dependence using a separable 2-dimensional covariance kernel, defined as the product of two (1-dimensional)
univariate covariance kernels. This introduces special symmetries and structure in the covariance matrix, which are
exploited in this package for fast and memory-efficient computations.

I developed snapKrig to support a project to interpolate weather data, but the methods underlying snapKrig
are applicable more generally. See also [[1](https://doi.org/10.7939/r3-g6qb-bq70)],
where I use product kernels to study directions of anisotropy in a non-stationary random fields, and
[[2](https://doi.org/10.1007/s11538-021-00899-z), [3](https://doi.org/10.1098/rsif.2020.0434)], where I apply it to
fit a covariance structure, and to speed up calculations of dispersal kernel convolutions.
